---
title: "Updated climate change projections released"
author: "Isaac Bain"
date: "2024-07-08"
categories: [code, climate, maps]
execute:
  warning: false
  error: false
  messages: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: left
    number-sections: true
editor: visual
lightbox: auto
title-block-banner: "#00000000"
title-block-banner-color: "rgba(255, 255, 255, 0.9)"
include-in-header:
  - text: |
      <style>
      #title-block-header.quarto-title-block.default .quarto-title-meta {
        color: rgba(255, 255, 255, 0.9);
      }
      .quarto-title-block .quarto-title-banner {
        height: 0; /* hide */
      }
      #title-block-header {
        background: 
          /* top, transparent black, faked with gradient */
          linear-gradient(
            rgba(0, 0, 0, 0.6),
            rgba(0, 0, 0, 0.2)
          ),
          /* bottom, image */ 
          url(./island-bay.JPG);
        background-size: cover;
        background-position-y: center;
        height: 300px;
        opacity: 0.7; /* image opacity, lower means lighter */
        z-index: -1;
      }
      </style>
---

## Climate change projections

[NIWA](https://niwa.co.nz/) have recently updated their national climate change projections for New Zealand, by downscaling six global models and three regional climate models through a series of machine learning approaches. This process was quite complex, with the model runs taking over a year to complete on a *literal* supercomputer, consuming a total of 12 million core hours. More details on the downscaling process can be found [here](https://niwa.co.nz/climate-and-weather/updated-national-climate-projections-aotearoa-new-zealand).

This work was publically funded by the Ministry for Business, Innovation and Employment (MBIE) and Ministry for the Environment (MfE), making the model outputs freely available for everyone to use. The climate projections data can be downloaded in NetCDF format from the new Climate Data Initiative website: <https://climatedata.environment.govt.nz/>.

## A quick primer on SSP and RCP

```{=html}
<blockquote class="instagram-media" data-instgrm-permalink="https://www.instagram.com/p/Bhv-iJtBX9_/?utm_source=ig_embed&amp;utm_campaign=loading" data-instgrm-version="14" style=" background:#FFF; border:0; border-radius:3px; box-shadow:0 0 1px 0 rgba(0,0,0,0.5),0 1px 10px 0 rgba(0,0,0,0.15); margin: 1px; max-width:540px; min-width:326px; padding:0; width:99.375%; width:-webkit-calc(100% - 2px); width:calc(100% - 2px);"><div style="padding:16px;"> <a href="https://www.instagram.com/p/Bhv-iJtBX9_/?utm_source=ig_embed&amp;utm_campaign=loading" style=" background:#FFFFFF; line-height:0; padding:0 0; text-align:center; text-decoration:none; width:100%;" target="_blank"> <div style=" display: flex; flex-direction: row; align-items: center;"> <div style="background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 40px; margin-right: 14px; width: 40px;"></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 100px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 60px;"></div></div></div><div style="padding: 19% 0;"></div> <div style="display:block; height:50px; margin:0 auto 12px; width:50px;"><svg width="50px" height="50px" viewBox="0 0 60 60" version="1.1" xmlns="https://www.w3.org/2000/svg" xmlns:xlink="https://www.w3.org/1999/xlink"><g stroke="none" stroke-width="1" fill="none" fill-rule="evenodd"><g transform="translate(-511.000000, -20.000000)" fill="#000000"><g><path d="M556.869,30.41 C554.814,30.41 553.148,32.076 553.148,34.131 C553.148,36.186 554.814,37.852 556.869,37.852 C558.924,37.852 560.59,36.186 560.59,34.131 C560.59,32.076 558.924,30.41 556.869,30.41 M541,60.657 C535.114,60.657 530.342,55.887 530.342,50 C530.342,44.114 535.114,39.342 541,39.342 C546.887,39.342 551.658,44.114 551.658,50 C551.658,55.887 546.887,60.657 541,60.657 M541,33.886 C532.1,33.886 524.886,41.1 524.886,50 C524.886,58.899 532.1,66.113 541,66.113 C549.9,66.113 557.115,58.899 557.115,50 C557.115,41.1 549.9,33.886 541,33.886 M565.378,62.101 C565.244,65.022 564.756,66.606 564.346,67.663 C563.803,69.06 563.154,70.057 562.106,71.106 C561.058,72.155 560.06,72.803 558.662,73.347 C557.607,73.757 556.021,74.244 553.102,74.378 C549.944,74.521 548.997,74.552 541,74.552 C533.003,74.552 532.056,74.521 528.898,74.378 C525.979,74.244 524.393,73.757 523.338,73.347 C521.94,72.803 520.942,72.155 519.894,71.106 C518.846,70.057 518.197,69.06 517.654,67.663 C517.244,66.606 516.755,65.022 516.623,62.101 C516.479,58.943 516.448,57.996 516.448,50 C516.448,42.003 516.479,41.056 516.623,37.899 C516.755,34.978 517.244,33.391 517.654,32.338 C518.197,30.938 518.846,29.942 519.894,28.894 C520.942,27.846 521.94,27.196 523.338,26.654 C524.393,26.244 525.979,25.756 528.898,25.623 C532.057,25.479 533.004,25.448 541,25.448 C548.997,25.448 549.943,25.479 553.102,25.623 C556.021,25.756 557.607,26.244 558.662,26.654 C560.06,27.196 561.058,27.846 562.106,28.894 C563.154,29.942 563.803,30.938 564.346,32.338 C564.756,33.391 565.244,34.978 565.378,37.899 C565.522,41.056 565.552,42.003 565.552,50 C565.552,57.996 565.522,58.943 565.378,62.101 M570.82,37.631 C570.674,34.438 570.167,32.258 569.425,30.349 C568.659,28.377 567.633,26.702 565.965,25.035 C564.297,23.368 562.623,22.342 560.652,21.575 C558.743,20.834 556.562,20.326 553.369,20.18 C550.169,20.033 549.148,20 541,20 C532.853,20 531.831,20.033 528.631,20.18 C525.438,20.326 523.257,20.834 521.349,21.575 C519.376,22.342 517.703,23.368 516.035,25.035 C514.368,26.702 513.342,28.377 512.574,30.349 C511.834,32.258 511.326,34.438 511.181,37.631 C511.035,40.831 511,41.851 511,50 C511,58.147 511.035,59.17 511.181,62.369 C511.326,65.562 511.834,67.743 512.574,69.651 C513.342,71.625 514.368,73.296 516.035,74.965 C517.703,76.634 519.376,77.658 521.349,78.425 C523.257,79.167 525.438,79.673 528.631,79.82 C531.831,79.965 532.853,80.001 541,80.001 C549.148,80.001 550.169,79.965 553.369,79.82 C556.562,79.673 558.743,79.167 560.652,78.425 C562.623,77.658 564.297,76.634 565.965,74.965 C567.633,73.296 568.659,71.625 569.425,69.651 C570.167,67.743 570.674,65.562 570.82,62.369 C570.966,59.17 571,58.147 571,50 C571,41.851 570.966,40.831 570.82,37.631"></path></g></g></g></svg></div><div style="padding-top: 8px;"> <div style=" color:#3897f0; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:550; line-height:18px;">View this post on Instagram</div></div><div style="padding: 12.5% 0;"></div> <div style="display: flex; flex-direction: row; margin-bottom: 14px; align-items: center;"><div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(0px) translateY(7px);"></div> <div style="background-color: #F4F4F4; height: 12.5px; transform: rotate(-45deg) translateX(3px) translateY(1px); width: 12.5px; flex-grow: 0; margin-right: 14px; margin-left: 2px;"></div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(9px) translateY(-18px);"></div></div><div style="margin-left: 8px;"> <div style=" background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 20px; width: 20px;"></div> <div style=" width: 0; height: 0; border-top: 2px solid transparent; border-left: 6px solid #f4f4f4; border-bottom: 2px solid transparent; transform: translateX(16px) translateY(-4px) rotate(30deg)"></div></div><div style="margin-left: auto;"> <div style=" width: 0px; border-top: 8px solid #F4F4F4; border-right: 8px solid transparent; transform: translateY(16px);"></div> <div style=" background-color: #F4F4F4; flex-grow: 0; height: 12px; width: 16px; transform: translateY(-4px);"></div> <div style=" width: 0; height: 0; border-top: 8px solid #F4F4F4; border-left: 8px solid transparent; transform: translateY(-4px) translateX(8px);"></div></div></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center; margin-bottom: 24px;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 224px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 144px;"></div></div></a><p style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; line-height:17px; margin-bottom:0; margin-top:8px; overflow:hidden; padding:8px 0 7px; text-align:center; text-overflow:ellipsis; white-space:nowrap;"><a href="https://www.instagram.com/p/Bhv-iJtBX9_/?utm_source=ig_embed&amp;utm_campaign=loading" style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:normal; line-height:17px; text-decoration:none;" target="_blank">A post shared by Carbon Brief (@carbonbrief)</a></p></div></blockquote> <script async src="//www.instagram.com/embed.js"></script>
```
### SSP (Shared Socioeconomic Pathways)

The SSPs are narratives that describe different global socio-economic futures. They are used in climate research to understand how social, economic, and environmental factors might evolve over the 21st century. There are five SSPs, each representing a different socio-economic development pathway:

1.  **SSP1 - Sustainability (Taking the Green Road)**: This pathway envisions a world making a shift towards sustainability, emphasizing inclusive development and respect for environmental limits.
2.  **SSP2 - Middle of the Road**: This pathway assumes moderate socio-economic development without major shifts in economic trends or technological advancements.
3.  **SSP3 - Regional Rivalry (A Rocky Road)**: This pathway describes a fragmented world with increasing nationalism, reduced international cooperation, and slow economic growth. Regional conflicts and challenges to sustainable development are more prevalent.
4.  **SSP4 - Inequality (A Road Divided)**: This pathway highlights a world with increasing inequalities, both within and between countries, leading to uneven socio-economic development.
5.  **SSP5 - Fossil-fueled Development (Taking the Highway)**: This pathway depicts rapid economic growth driven by a reliance on fossil fuels and energy-intensive lifestyles.

### RCP (Representative Concentration Pathways)

The RCPs are scenarios that include projections of future greenhouse gas (GHG) concentrations based on different trajectories of GHG emissions. Each RCP represents a different level of radiative forcing (the change in energy in the atmosphere due to GHGs) by the year 2100, for example:

1.  **RCP2.6**: A scenario with low GHG emissions leading to a peak radiative forcing of around 2.6 W/m² before 2100, followed by a decline.
2.  **RCP4.5**: A scenario with intermediate GHG emissions leading to a stabilization of radiative forcing at around 4.5 W/m² by 2100.
3.  **RCP6.0**: Another intermediate scenario with radiative forcing reaching 6.0 W/m² by 2100.
4.  **RCP8.5**: A high GHG emissions scenario leading to radiative forcing reaching 8.5 W/m² by 2100.

## Climate variables

NIWA has modelled the projected changes for the periods 2021-2040, 2041-2060, and 2080-2099 under three RCP-SSP scenarios, using two baseline periods (1986-2005 and 1995-2014) for 24 climate variables. These variables include:

-   Temperature
-   Growing degree days
-   Rainfall
-   Potential Evapotranspiration Deficit
-   Wind
-   Humidity
-   Solar radiation

For the full list see [here](https://climatedata.environment.govt.nz/variables.html).

## Data visualisation

This post was mainly to draw your attention to the myriad of new datasets that are becoming available. However, I've selected one dataset to visualise here to give you a little taste:

\- How total **rainfall** is estimated to change under an **SSP-3 RCP-7.0** scenario **compared to a 1986-2005** baseline.

The data is in NetCDF format, which is a common format for storing multidimensional scientific data. We'll use the `tidync` package to read the NetCDF files and the `ggplot2` package to create the visualisations. [`tidync`](https://docs.ropensci.org/tidync/) brings tidyverse concepts to NetCDF files.

Let's start by mapping the baseline period, from which the projected change is compared.

```{r fig.height=7, fig.width=6}
# Load necessary libraries
library(tidyverse)
library(tidync)
library(ggplot2)
library(gganimate)
library(viridis)
library(ggtext)

# Set path and filenames
ncpath <- "data/PR_epsg_4272/"
ncvar <- "PR_"
baseline_nc_file <- "historical_MMM_CCAM_base_bp1986-2005_ANN_NZ5km"

baseline_nc_fname <- paste(ncpath, ncvar, baseline_nc_file, ".nc", sep="")
  
baseline_data <- tidync(baseline_nc_fname) |>  
    hyper_tibble()

ggplot(baseline_data, aes(x = longitude, y = latitude, fill = log(PR))) +
  geom_tile(width = 0.07, height = 0.07) +
  coord_map() +
  scale_fill_distiller(
    palette = "Blues",
    direction = 1,
    labels = function(x) paste0(round(exp(x), 0), " mm")) +
    # breaks = log(c(300, 1000, 3000, 5000, 10000, 14500)),
    # labels = c("300 mm", "1000 mm", "3000 mm", "5000 mm", "10000 mm", "14500 mm")) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b"
  )) +
  labs(
    title = "Baseline total rainfall, 1986-2005",
    fill = "Annual rainfall",
    x = "Longitude", 
    y = "Latitude"
  )
```

This code snippet provides a starting point for visualising the baseline period rainfall. You can adjust the path to the NetCDF file and further customise the plot as needed.

Now let's map the projected change in rainfall for the same region under the SSP-3 RCP-7.0 scenario.

```{r}
#| fig.width: 6
#| fig.height: 7
#| out-width: 80%

ssp <- "ssp370_"
ncfiles <- c(
  "MMM_CCAM_change_fp2021-2040_bp1986-2005_ANN_NZ5km",
  "MMM_CCAM_change_fp2041-2060_bp1986-2005_ANN_NZ5km",
  "MMM_CCAM_change_fp2080-2099_bp1986-2005_ANN_NZ5km"
)

# Initialize an empty data frame to store the combined data
all_data <- tibble()

# Function to extract year range from file name
extract_years <- function(ncname) {
  str_extract(ncname, "fp\\d{4}-\\d{4}") %>% 
    str_replace("fp", "")
}

# Iterate over the file names and extract data
for (ncname in ncfiles) {
  ncfname <- paste(ncpath, ncvar, ssp, ncname, ".nc", sep="")
  
  tdata <- tidync(ncfname) %>% 
    hyper_tibble() %>%
    mutate(period = extract_years(ncname))
  
  all_data <- bind_rows(all_data, tdata)
}

# Ensure longitude and latitude are numeric
all_data <- all_data %>%
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude))

# Determine the overall range for the PR variable
pr_range <- range(all_data$PR, na.rm = TRUE)

# Plot and animate the data
p <- ggplot(all_data, aes(x = longitude, y = latitude, fill = PR)) +
  geom_tile(width = 0.07, height = 0.07) +
  coord_map() +
  scale_fill_distiller(
    palette = "RdBu",
    limits = pr_range,
    direction = 1,
    labels = function(x) paste0(x, "%")) +
  labs(
    title = "Projected change in total rainfall under SSP-3 RCP-7.0",
    subtitle = 'Baseline of 1986–2005 compared to: {closest_state}',
    fill = "Change from \nbaseline",
    x = "Longitude", 
    y = "Latitude"
  ) +
  annotate(
    "text", x = 168.546875, y = -42.836828, label = "The west gets \n", hjust = 0.5, color = "black", size = rel(3)
  ) +
  annotate(
    "text", x = 168.546875, y = -43.036828, label = "wetter", hjust = 0.5, color = "#2257a0", size = rel(3.5), face = "bold"
  ) +
  annotate(
    "text", x = 177.873145, y = -40.89288, label = "The east gets \n", hjust = 0.5, color = "black", size = rel(3)
  ) +
  annotate(
    "text", x = 177.873145, y = -41.09288, label = "drier", hjust = 0.5, color = "#a61729", size = rel(3.5), face = "bold"
  ) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "right") + 
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b")) +
  theme(plot.subtitle = element_text(color = "#222222")) +
  transition_states(period, transition_length = 2, state_length = 1)

animate(p,
        nframes = 100,
        fps = 20,
        res = 400,
        renderer = gifski_renderer())

```

## Conclusion

Understanding future climate change impacts is crucial for planning and adaptation. By leveraging advanced climate models and high-performance computing, NIWA's projections offer valuable insights into how New Zealand's climate might change under different scenarios.

As these datasets become increasingly accessible, they empower stakeholders to develop more resilient strategies to cope with the changing climate. Other users will undoubtedly find even more valuable applications for these data, enhancing their utility and impact across various sectors.
