---
title: "Arrowtown air quality"
author: "Isaac Bain"
date: "2024-06-02"
categories: [code, air quality, animation]
execute:
  warning: false
  error: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
---

Most of the time, New Zealand enjoys good air quality, allowing us to step outside and breathe a breath of fresh air, knowing it's safe. However at certain times and places, our air is polluted to such levels that it causes us harm, or in the worst cases, premature death. Here I access air quality data from LAWA and explore it for one site known for severe winter air pollution.

## Libraries

```{r}
library(tidyverse) # provides ggplot, dplyr, etc
library(janitor) # tidy up messy variable names
library(gganimate) # provides animations
library(geomtextpath) # makes labels curved on polar graph
library(leaflet) # for interactive mapping
```

## Import data

[Land, Air, Water Aotearoa (LAWA)](https://lawa.org.nz/) is a collaboration between New Zealand's 16 regional councils and unitary authorities, the Cawthron Institute, the Ministry for the Environment, the Department of Conservation, and Stats NZ. LAWA provides state of the environment monitoring data for various [environmental domains](https://lawa.org.nz/explore-data), including; swimming water quality, estuary health, groundwater quality, lake quality, land cover, river quality, water quantity, and of particular interest here, air quality.

LAWA makes their data freely available at <https://lawa.org.nz/download-data>. From this source I downloaded an Excel sheet of air quality data for the period 2016 to 2022. Currently, the available parameters are PM~10~ and PM~2.5~.

Particulate matter (PM) can have adverse effects on human health and the environment. The numbers 10 and 2.5 refer to the diameter of the particles in micrometers (µm).

::: {.callout-note collapse="false"}
## What are the human health impacts of PM in New Zealand?

Hales et al. (2021) did a retrospective cohort epidemiological study of the association between PM~2.5~, NO~2~ and mortality and morbidity for 2.2m New Zealanders. They found that exposure to PM~2.5~ per 10 μg/m^3^ increased the risk of death by 11% (95% CI 7-15%) and increased the risk of ischaemic heart disease by 29% (95% CI 23-35%).

Hales, S., Atkinson, J., Metcalfe, J., Kuschel, G., & Woodward, A. (2021). Long term exposure to air pollution, mortality and morbidity in New Zealand: cohort study. *Science of The Total Environment, 801*, 149660.
:::

```{r}
dat <- readxl::read_xlsx("data/airqualitydownloaddata_2016-2022.xlsx",
                         sheet = "Air Quality Data") |> # read in data
  clean_names() |> # tidy up messy names
  mutate(year = year(sample_date)) |> # extract year as a new variable
  filter(!is.na(sample_date) & !is.na(concentration)) # filter out NAs
```

## What are the worst sites?

Air quality is known to be worst in more southern areas, places where home heating via wood burning or coal is more prevalent, particularly during the colder months of the year.

But let's check some of these assumptions by extracting the worst 5 sites for highest recorded PM~10~ values (@tbl-top5).


```{r}
#| label: tbl-top5
#| tbl-cap: "Top 5 sites with highest PM~10~ concentrations"

dat |> 
  filter(indicator == "PM10") |> # just look at PM10
  group_by(site_name) |> 
  summarize(max_concentration = max(concentration, na.rm = TRUE)) |> # get max conc. per site
  arrange(desc(max_concentration)) |> # sort descending
  top_n(5, max_concentration) |> # get top 5 sites
  knitr::kable()
```

Indeed most of these locations are in South Island, with the exception of Lower Hutt, which is at the southern end of the North Island.

```{r}
processed_data <- dat %>% 
  filter(indicator == "PM10") %>% # just look at PM10
  group_by(site_name) %>% 
  summarise(
    max_concentration = max(concentration, na.rm = TRUE), 
    latitude = first(latitude),
    longitude = first(longitude)
  ) %>% # get max conc. per site
  arrange(desc(max_concentration)) %>% # sort descending
  top_n(5, max_concentration) %>% 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Create a color palette
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = processed_data$max_concentration, # Set the domain to the range of max_concentration
  na.color = "transparent",
  reverse = TRUE
)

# Create a Leaflet map
leaflet(processed_data) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = ~sqrt(max_concentration), # Scale the radius proportional to max_concentration
    color = ~pal(max_concentration),
    fillOpacity = 1,
    stroke = 0,
    label = ~paste0(site_name, ", ", "Max PM10: ", max_concentration)
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = processed_data$max_concentration, # Pass the actual max_concentration values
    title = "Max PM10 Concentration",
    opacity = 0.7,
    labFormat = labelFormat(transform = function(x) sort(x, decreasing = TRUE))
  )
```

## Static plot of seasonal air quality

Now let's check the assumption that air quality is worse during the colder months of the year. Here, I plot daily PM~10~ concentrations over 2022 for one site (Arrowtown) (@fig-staticplot). The data shows a significant spike during the winter months, with a number of exceedances of the WHO guideline of 45 µg/m³.

```{r}
#| label: fig-staticplot
#| fig-cap: "PM~10~ daily concentrations at Arrowtown, 2022. Dashed line shows WHO guideline value."

dat_filtered <- dat |> 
  filter(site_name == "Arrowtown") |> 
  filter(indicator == "PM10") |> 
  filter(year == 2022) |>
  arrange(sample_date) |> 
  mutate(sample_date = as.Date(sample_date))

# Calculate the breaks for y-axis labels
n_breaks <- 5
my_breaks <- scales::breaks_extended(n = n_breaks)(range(dat_filtered$concentration))
my_breaks <- my_breaks[my_breaks > 0]

# Create a data frame for the tick marks
tick_data <- data.frame(
  x = as.Date("2022-01-01"),
  xend = as.Date("2022-01-03"),
  y = my_breaks,
  yend = my_breaks
)

# plot
ggplot(dat_filtered) +
  geom_hline(yintercept = 45, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_line(aes(x = sample_date, y = concentration),
            color = "#D72638",
            size = 0.5) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = bquote("Daily PM"[10] * " concentration (µg/m"^3 * ") in Arrowtown, 2022"),
    x = "",
    y = bquote("Daily PM"[10] * " concentration (µg/m"^3 * ")")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, 
                              colour = "#420039"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(
      color = "#12355B",
      face = "bold",
      size = 14,
      vjust = -0.5
    )
  ) +
  annotate(
    'text',
    x = as.Date("2022-01-03"),
    y = my_breaks,
    label = round(my_breaks, 2),
    angle = 0,
    hjust = -0.1,
    vjust = 0.5
  ) +
  geom_segment(data = tick_data,
               aes(
                 x = x,
                 xend = xend,
                 y = y,
                 yend = yend
               ),
               color = "black")
```

## Make an animated plot

Let's enhance our analysis by creating an animated polar chart that reveals the daily PM~10~ concentrations over time (@fig-animation). This approach effectively highlights the cyclical nature of seasonal air quality. By animating the data, we can observe the progression of PM~10~ levels throughout the year, providing a clear visual representation of how air quality deteriorates during the winter months.

```{r}
#| label: fig-animation
#| fig-cap: "Animated polar chart of PM~10~ daily concentrations at Arrowtown, 2022. Dashed circle shows WHO guideline value. Dashed lines show seasons."

# Define the start dates for each season according to New Zealand definitions
season_start_dates <- as.Date(c("2022-12-01", "2022-03-01", "2022-06-01", "2022-09-01"))

# Create the plot with seasonal division lines and y-axis labels inside the chart
p <- ggplot(dat_filtered) +
  geom_hline(yintercept = 45, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_line(aes(x = sample_date, y = concentration),
            color = "#D72638",
            size = 0.5) +
  geom_vline(
    xintercept = as.numeric(season_start_dates),
    linetype = "dashed",
    color = "black",
    alpha = 0.3
  ) +
  geom_point(aes(x = sample_date, y = concentration), color = "#420039", size = 2) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  coord_curvedpolar() +
  labs(
    title = bquote("Daily PM"[10] * " concentration (µg/m"^3 * ") in Arrowtown, 2022"),
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, 
                              colour = "#420039"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(
      color = "#12355B",
      face = "bold",
      size = 14,
      vjust = -0.5
    )
  ) +
  annotate(
    'text',
    x = as.Date("2022-01-03"),
    y = my_breaks,
    label = round(my_breaks, 2),
    angle = 0,
    hjust = -0.1,
    vjust = 0.5
  ) +
  geom_segment(data = tick_data,
               aes(
                 x = x,
                 xend = xend,
                 y = y,
                 yend = yend
               ),
               color = "black") +
  transition_reveal(sample_date) +
  ease_aes('linear')  # Easing to make the animation smoother

# Render the animation and save as GIF
anim <- animate(p, nframes = 100, fps = 10, renderer = gifski_renderer())

anim
```

## Make 4 animated plots

Now let's check out whether other high PM~10~ sites also have similar seasonal dynamics in air quality (we'll also use a different year to make sure) (@fig-top4).

```{r}
#| label: fig-top4
#| fig-cap: "Animated polar chart of PM~10~ daily concentrations at Arrowtown, Invercargill, Lower Hutt, and Milton, 2019. Dashed circle shows WHO guideline value. Note the y-axes differ between plots."

top_4 <- dat |> 
  filter(indicator == "PM10") |> # just look at PM10
  group_by(site_name) |> 
  summarize(max_concentration = max(concentration, na.rm = TRUE)) |> # get max conc. per site
  arrange(desc(max_concentration)) |> # sort descending
  top_n(4, max_concentration)

p2 <- dat |> 
  filter(site_name %in% top_4$site_name) |> 
  filter(indicator == "PM10") |> 
  filter(year == 2019) |>
  arrange(sample_date) |> 
  mutate(sample_date = as.Date(sample_date)) |> 
  ggplot() +
    geom_hline(yintercept = 45, linetype = "dashed", color = "red", alpha = 0.7) +
    geom_line(aes(x = sample_date, y = concentration), color = "#D72638", size = 0.5) +
    geom_point(aes(x = sample_date, y = concentration), color = "#420039", size = 2) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    coord_curvedpolar() +
    theme_minimal() +
    labs(x = "", 
         y = bquote("Daily PM"[10] * " concentration (µg/m"^3 * ")")) +
    transition_reveal(sample_date) +
    ease_aes('linear') +
    facet_wrap(~site_name, scales = "free_y")

# Render the animation and save as GIF
anim2 <- animate(p2, nframes = 100, fps = 10, renderer = gifski_renderer())

anim2
```

Some interesting results here.[^1] Firstly, Arrowtown demonstrates similar seasonal dynamics to 2022, with low concentrations in summer and a number of exceedances during winter. Invercargill also shows similar seasonality, albeit at lower concentrations.

[^1]: Note the y-axes differ between each site. This is due to the large magnitude in differences. My goal here was to explore the relative seasonality between each site, so having free axes makes sense. If the goal was instead to explore relative magnitudes between each site, I could have simply set `facet_wrap(~site_name, scales = "fixed")`.

Lower Hutt remains well below the guideline all year-round, but exhibits reversed seasonality, with higher concentrations in summer. It's worth noting that Lower Hutt is included in the top four due to a single large spike in 2022, so it might be reasonable to exclude this site from further analysis.

Milton is particularly interesting as it has numerous exceedances during winter but appears not to monitor air quality during summer (this trend is true in other years as well).
