---
title: "Map of South Island remote huts"
author: "Isaac Bain"
date: "2024-07-2"
categories: [code, maps]
execute:
  warning: false
  error: false
  messages: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: left
    number-sections: true
editor: visual
lightbox: auto
draft: true
title-block-banner: "#00000000"
title-block-banner-color: "rgba(255, 255, 255, 0.9)"
include-in-header:
  - text: |
      <style>
      #title-block-header.quarto-title-block.default .quarto-title-meta {
        color: rgba(255, 255, 255, 0.9);
      }
      .quarto-title-block .quarto-title-banner {
        height: 0; /* hide */
      }
      #title-block-header {
        background: 
          /* top, transparent black, faked with gradient */
          linear-gradient(
            rgba(0, 0, 0, 0.6),
            rgba(0, 0, 0, 0.2)
          ),
          /* bottom, image */ 
          url(./image.jpeg);
        background-size: cover;
        background-position-y: center;
        height: 300px;
        opacity: 0.7; /* image opacity, lower means lighter */
        z-index: -1;
      }
      </style>
---

## Overview

The Permolat Trust maintains a unique and iconic network of around 70 remote back-country huts and bivouacs in Central Westland. These huts were in a dilapidated state and under threat of disappearing due to lack of use and funding - however, the reason they weren't being used was because people of the difficulty in accessing them and obtaining up-to-date route information!

## Data sources

The Trust has created the [Remote Huts](https://www.remotehuts.co.nz/) website which is a fantastic resource for obtaining key details about each of their huts. Each individual hut page features a topographic map showing the hut and its surrounds. However, the site lacked an overview map displaying the relative location of all their huts - which would be useful for trip planning and discovering new huts to visit.

After confirming with the site maintainer they did not have a database of their hut listings in the back-end, I scraped all the necessary details into a Google Sheet.

Libraries

```{r}
#| label: libraries

library(tidyverse)
library(googlesheets4)
library(mapview)
library(sf)
library(leaflet)
library(htmltools)
library(ggmap)
```

Import data

```{r}
#| label: import-data
#| cache: true

gs4_deauth()

df <- read_sheet("https://docs.google.com/spreadsheets/d/1bh7kHtU--3jA8gb02yBaqBhp6PH5mdevNWOjkc41EXg/edit#gid=0") |>
  st_as_sf(coords = c("X", "Y"), crs = 4326)

```

## Mapping

### Static map

The static map provides a visual overview of the Permolat Trust's network of huts across the South Island. Using the `ggmap` package, a base layer of the South Island was retrieved and plotted with the hut locations overlaid.

```{r}
#| label: static-map

# bounding box for South Island
si <- c(left = 166.25, bottom = -47.5, right = 174.5, top = -40.0)

# retrieve base layer
map <- get_stadiamap(si, zoom = 8, maptype = "stamen_terrain_background") 

# Define a function to fix the bbox to be in EPSG:3857
# https://stackoverflow.com/questions/47749078/how-to-put-a-geom-sf-produced-map-on-top-of-a-ggmap-produced-raster
ggmap_bbox <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))

  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))

  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}

# Use the function:
map <- ggmap_bbox(map)

# create the static map
ggmap(map) +
  coord_sf(crs = st_crs(3857)) + # force the ggplot2 map to be in 3857
  geom_sf(data = df |> st_transform(3857),
          inherit.aes = FALSE, 
          aes(),
          color = "blue",
          size = 1.5)
```

### Interactive map

The interactive map offers a dynamic way to explore the Permolat Trust's huts across the South Island. It includes several user-friendly features:

-   **Hover to View Details:** Users can hover over any point to see the hut's name and a photo, making it easy to identify and learn more about each hut.

-   **Clickable Points:** Clicking on any point takes users directly to the Remote Huts website for detailed information about the hut.

-   **Customizable Base Maps:** Users can switch between different base maps, including terrain, Topographic 1:50k and 1:250k, and satellite views. This functionality, available via a control in the upper right corner, allows users to choose the most useful map type for their needs.

The interactive map is created using the `leaflet` package, which provides powerful tools for building interactive maps. Here’s a breakdown of the key steps:

1.  **Popup Content:** The hover popups display the hut's name and image, generated using HTML content.

2.  **Click Event:** JavaScript is used to add click events to each point, linking to the corresponding hut page on the Remote Huts website.

3.  **Base Layers:** Various base map layers are added using `addProviderTiles` and `addTiles`, giving users multiple perspectives of the terrain.

```{r}
#| label: html-map
#| column: screen
# column: screen-inset-shaded

# create hover popup spec
content <- paste0("<div style='text-align: center;'><b>", df$Name, "</b></div>", "<img src = '", df$Image, "', width = '175'>")

# goto url on click
jsCode <- paste0('
 function(el, x, data) {
  var markers = document.getElementsByClassName("leaflet-interactive");
  for(var i = 0; i < markers.length; i++){
    (function(){
      var v = data.URL[i];
      markers[i].addEventListener("click", function() { window.open(v); }, false);
    }()); 
  }
 }
')

# setup xyz tiles
topo50_tiles <- paste0(
  "https://tiles-cdn.koordinates.com/services;key=", Sys.getenv("koordinates_tile_api"), "/tiles/v4/layer=50767/EPSG:3857/{z}/{x}/{y}.png"
  )

topo250_tiles <- paste0(
  "https://tiles-cdn.koordinates.com/services;key=", Sys.getenv("koordinates_tile_api"), "/tiles/v4/layer=50798/EPSG:3857/{z}/{x}/{y}.png"
)

# create leaflet map
map <- 
  leaflet(df) %>%
  addMapPane(name = "circles", zIndex = 410) %>%
  addCircleMarkers(
    color = "black",
    fillColor = "blue",
    radius = 5,
    fillOpacity = 0.75, 
    stroke = T, 
    weight = 1,
    opacity = 1,
    label = lapply(content, htmltools::HTML),
  ) %>%
  addProviderTiles(providers$CartoDB.Voyager, group = "Terrain") %>%
  addTiles(topo50_tiles, group = "Topo50") %>%
  addTiles(topo250_tiles, group = "Topo250") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  addLayersControl(
    baseGroups = c("Terrain", "Topo250", "Topo50", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  htmlwidgets::onRender(jsCode, data = df) 

map
```

## Conclusion

Finding Remote Huts is now much easier with the overview map of the South Island. I have also uploaded a copy of my map to the Remote Huts website, allowing users to explore by A-Z, by river catchment, and now [by map](https://www.remotehuts.co.nz/by-map.html)!

This is a great result for backcountry users and the Permolat Trust, which now has a web-map integrated into their existing website infrastructure.
