---
title: "Bivariate colour scales"
subtitle: "Mapping livestock density and water quality"
author: "Isaac Bain"
date: "2024-07-09"
categories: [code, water quality, maps]
execute:
  warning: false
  error: false
  messages: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: left
    number-sections: true
editor: visual
lightbox: auto
draft: true
title-block-banner: "#00000000"
title-block-banner-color: "rgba(255, 255, 255, 0.9)"
include-in-header:
  - text: |
      <style>
      #title-block-header.quarto-title-block.default .quarto-title-meta {
        color: rgba(255, 255, 255, 0.9);
      }
      .quarto-title-block .quarto-title-banner {
        height: 0; /* hide */
      }
      #title-block-header {
        background: 
          /* top, transparent black, faked with gradient */
          linear-gradient(
            rgba(0, 0, 0, 0.6),
            rgba(0, 0, 0, 0.2)
          ),
          /* bottom, image */ 
          url(./cows.jpg);
        background-size: cover;
        background-position-y: center;
        height: 300px;
        opacity: 0.7; /* image opacity, lower means lighter */
        z-index: -1;
      }
      </style>
---

## Overview

We all want river that are safe to swim in, clean water to drink, and healthy freshwater ecosystems. But many areas in New Zealand are not living up to our communities' expectations and require improvement.[^1] Knowing where to target efforts to improve water quality is crucial, and a solid understanding of pressures is needed.

[^1]: <https://environment.govt.nz/publications/our-freshwater-2023/>

Land use intensity is consistently identified as a key pressure on water quality in New Zealand.[^2] In both process-based and statistical models predicting water quality it is often the most, or one of the most, significant predictors of *E. coli* and nitrogen concentrations.

[^2]: Scott T. Larned, Jonathan Moores, Jenni Gadd, Brenda Baillie & Marc Schallenberg (2020) Evidence for the effects of land use on freshwater ecosystems in New Zealand, New Zealand Journal of Marine and Freshwater Research, 54:3, 551-591, DOI: 10.1080/00288330.2019.1695634

For example, in NIWA's most recent random forest modelling of river water quality[^3], the top ten most important predictors of nitrate + nitrite-nitrogen (NNN) and *E. coli* were:

[^3]: <https://environment.govt.nz/assets/publications/spatial-modelling-river-quality.pdf>

| Predictor rank | NNN predictors                   | *E. coli* predictors             |
|------------------|---------------------------|---------------------------|
| 1              | Stock density                    | Catchment elevation              |
| 2              | Dairy density                    | Rainfall variability             |
| 3              | Beef density                     | Stock density                    |
| 4              | Intensive agriculture land cover | Bare land cover                  |
| 5              | Urban land cover                 | Intensive agriculture land cover |
| 6              | Catchment slope                  | Slope                            |
| 7              | Temperature Minimum              | Urban land cover                 |
| 8              | Rainfall variability             | Exotic forest land cover         |
| 9              | Number of warm days              | Local elevation                  |
| 10             | Catchment elevation              | Catchment phosphorus             |

\ 
In this analysis, I'm interested in finding out which areas of New Zealand have relatively high or low levels of contamination given relativley high or low livestock density. Which lends itself well to visualising using a bivariate colour scale.

## Bivariate maps

Bivariate colour scale maps are thematic maps that use two different colour scales to simultaneously display two variables on the same map. This approach allows for the visualisation of the relationship between the two variables across a geographic space. Each colour on the map represents a unique combination of the two variables' values.

Typically, one colour gradient is represented on the x-axis and the other on the y-axis. The intersection of these gradients results in a grid of colours, each representing a different combination of the two variables' values.

```{r}
#| label: libraries
#| output: false
#| code-summary: "Show libraries code"

library(tidyverse)
library(sf)
library(mapview)
library(koordinatr)
library(biscale)
library(cowplot)
library(patchwork)

```

## Data sources

For this example, river water quality data modelled for every stream segment in the digital river network - this was done by NIWA on behalf of MfE.

```{r}
#| label: import
#| cache: true
#| output: false
#| code-summary: "Show data sources code"

# water quality
# ecoli_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109886")
# saveRDS(ecoli_raw, "data/ecoli_raw.rds")
ecoli_raw <- readRDS("data/ecoli_raw.rds")

ecoli <- ecoli_raw |>
  filter(mesrmnt == "Median") |>
  select(nzsgmnt,
    median_ecoli = value,
    strm_rd
  )

rm(ecoli_raw)

# nitrogen_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109888")
# saveRDS(nitrogen_raw, "data/nitrogen_raw.rds")
nitrogen_raw <- readRDS("data/nitrogen_raw.rds")

nitrogen <- nitrogen_raw |>
  filter(measr_b == "NNN") |>
  select(nzsgmnt,
    median_nitrogen = value
  )

rm(nitrogen_raw)

# load predictors
load("data/REC2.4_lcdb5_predictors.RData")
rm(ThesePredictors)

ecoli |>
  left_join(nitrogen |> st_drop_geometry(), by = join_by(nzsgmnt)) |>
  left_join(REC2.4_predictors, by = c("nzsgmnt" = "nzsegment")) |>
  select(
    nzsgmnt,
    median_ecoli,
    median_nitrogen,
    PropDairy_2017,
    PropSheep_2017,
    SUDensityTotal_2017,
    StreamOrder
  ) |>
  filter(StreamOrder > 1) -> dat
```

## Livestock vs water quality

Let's start by looking at the relationship between dairy density and median *E. coli* levels. We can see that there is a positive relationship between the two variables - areas with higher dairy density tend to have higher *E. coli* levels (@fig-scatter-plot). 

However, the relationship is not perfect - there are areas with high dairy density and low *E. coli* levels, and vice versa. This is where bivariate maps come in - they allow us to visualise the relationship between two variables across a geographic space.

```{r}
#| label: fig-scatter-plot
#| fig-cap: "Scatter plot of dairy density and median E. coli levels."

dat |> 
  ggplot(aes(PropDairy_2017, median_ecoli)) +
  geom_point(alpha = 0.01, size = 1) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(x = "Dairy density", y = "Median E. coli")
```
## Univariate maps

Let's take a look at our data sources using traditional maps which just use a single colour scale to display a single variable (@fig-uni-maps). They are not bad - but it's pretty hard to imagine where has high dairy density and low *E. coli* (and vice versa) right? That's where bivariate maps come in.

```{r}
#| label: fig-uni-maps
#| fig-cap: "Univariate maps of median E. coli levels and dairy density."
#| code-summary: "Show univariate code"

dat |>
  arrange(median_ecoli) |>
  ggplot() +
  geom_sf(aes(colour = median_ecoli)) +
  # scale_colour_viridis_c() +
  scale_colour_gradient(low = "#f2f2f2", high = "#0295ff") +
  theme_void() +
  labs(title = "E. coli concentrations") -> ecoli_map

dat |>
  ggplot() +
  geom_sf(aes(colour = PropDairy_2017)) +
  scale_colour_gradient(low = "#f2f2f2", high = "#ff9400") +
  # scale_fill_viridis_c() +
  theme_void() +
  labs(title = "Dairy density") -> dairy_dens_map

ecoli_map + dairy_dens_map
```

## Bivariate colour scale maps

```{r}
#| label: spatial-join
#| code-summary: "Show prep code"

custom_pal4 <- c(
  "1-1" = "#f2f2f2", # low x, low y #ffffff
  "2-1" = "#b9ddff",
  "3-1" = "#64bbff",
  "4-1" = "#0295ff", # high x, low y
  "1-2" = "#fedcb7",
  "2-2" = "#b8b8b8",
  "3-2" = "#6691bb",
  "4-2" = "#0069ba",
  "1-3" = "#ffb869",
  "2-3" = "#b89168",
  "3-3" = "#676767",
  "4-3" = "#003867",
  "1-4" = "#ff9400", # low x, high y
  "2-4" = "#bb6800",
  "3-4" = "#693800",
  "4-4" = "#000000" # high x, high y
)

# bi_pal(custom_pal4, dim = 4, flip_axes = TRUE)

finalplot <- function(data, xlab = "Higher X", ylab = "Higher Y") {
  map <- ggplot() +
    geom_sf(
      data = data, mapping = aes(colour = bi_class), linewidth = 0.5, show.legend = FALSE,
      # color = "white"
    ) +
    bi_scale_color(pal = custom_pal4, dim = 4) +
    bi_theme()

  legend <- bi_legend(
    pal = custom_pal4,
    dim = 4,
    xlab = xlab,
    ylab = ylab,
    size = 7,
    pad_width = 0.5
  )

  finalPlot <- ggdraw() +
    draw_plot(map, 0, 0, 1, 1) +
    draw_plot(legend, 0.2, .65, 0.2, 0.2, 1.8)

  return(finalPlot)
}
  
```

### Total livestock density

::: panel-tabset
## E. coli

```{r, dpi=300}
#| label: fig-bi-maps-total-ecoli

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = SUDensityTotal_2017,
  style = "fisher", dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher stock"
  ) -> finalPlotTotalEcoli

finalPlotTotalEcoli
```

## Nitrogen

```{r, dpi=300}
#| label: fig-bi-maps-total-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = SUDensityTotal_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher stock"
  ) -> finalPlotTotalNitrogen

finalPlotTotalNitrogen
```
:::

### Sheep density

::: panel-tabset

## E. coli

```{r}
#| label: fig-bi-maps-sheep-ecoli

bi_class(
  dat,
  x = median_ecoli,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher sheep"
  ) -> finalPlotSheepEcoli

finalPlotSheepEcoli
```

## Nitrogen

```{r}
#| label: fig-bi-maps-sheep-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher sheep"
  ) -> finalPlotSheepNitrogen

finalPlotSheepNitrogen
```
:::


### Dairy density

::: panel-tabset

## E. coli

```{r}
#| label: fig-bi-maps-dairy-ecoli

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher dairy"
  ) -> finalPlotDairyEcoli

finalPlotDairyEcoli
```

## Nitrogen

```{r}
#| label: fig-bi-maps-dairy-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher dairy"
  ) -> finalPlotDairyNitrogen

finalPlotDairyNitrogen
```
:::

## Conclusion
