---
title: "Bivariate colour scales"
subtitle: "Mapping livestock density and water quality"
author: "Isaac Bain"
date: "2024-07-09"
categories: [code, water quality, maps]
execute:
  warning: false
  error: false
  messages: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: left
    number-sections: true
editor: visual
lightbox: auto
draft: true
title-block-banner: "#00000000"
title-block-banner-color: "rgba(255, 255, 255, 0.9)"
include-in-header:
  - text: |
      <style>
      #title-block-header.quarto-title-block.default .quarto-title-meta {
        color: rgba(255, 255, 255, 0.9);
      }
      .quarto-title-block .quarto-title-banner {
        height: 0; /* hide */
      }
      #title-block-header {
        background: 
          /* top, transparent black, faked with gradient */
          linear-gradient(
            rgba(0, 0, 0, 0.6),
            rgba(0, 0, 0, 0.2)
          ),
          /* bottom, image */ 
          url(./cows.jpg);
        background-size: cover;
        background-position-y: center;
        height: 300px;
        opacity: 0.7; /* image opacity, lower means lighter */
        z-index: -1;
      }
      </style>
---

## Overview

We all want river that are safe to swim in, clean water to drink, and healthy freshwater ecosystems. But many areas in New Zealand are not living up to our communities' expectations and require improvement.[^1] Knowing where to target efforts to improve water quality is crucial, and a solid understanding of pressures is needed.

[^1]: <https://environment.govt.nz/publications/our-freshwater-2023/>

In order for management actions to be effective, they should be based on predictive relationships between land use, stressor levels in receiving waters, and the ecological and cultural responses to these stressors.[^2] Land use intensity is consistently identified as a key pressure on water quality in New Zealand. In both process-based and statistical models predicting water quality it is often the most, or one of the most, significant predictors of *E. coli* and nitrogen concentrations.

[^2]: Scott T. Larned, Jonathan Moores, Jenni Gadd, Brenda Baillie & Marc Schallenberg (2020) Evidence for the effects of land use on freshwater ecosystems in New Zealand, New Zealand Journal of Marine and Freshwater Research, 54:3, 551-591, DOI: 10.1080/00288330.2019.1695634

For example, in NIWA's most recent random forest modelling of river water quality[^3], the top ten most important predictors of nitrate + nitrite-nitrogen (NNN) and *E. coli* were:

[^3]: <https://environment.govt.nz/assets/publications/spatial-modelling-river-quality.pdf>

| Predictor rank | NNN predictors                   | *E. coli* predictors             |
|----------------|----------------------------------|----------------------------------|
| 1              | Stock density                    | Catchment elevation              |
| 2              | Dairy density                    | Rainfall variability             |
| 3              | Beef density                     | Stock density                    |
| 4              | Intensive agriculture land cover | Bare land cover                  |
| 5              | Urban land cover                 | Intensive agriculture land cover |
| 6              | Catchment slope                  | Slope                            |
| 7              | Temperature Minimum              | Urban land cover                 |
| 8              | Rainfall variability             | Exotic forest land cover         |
| 9              | Number of warm days              | Local elevation                  |
| 10             | Catchment elevation              | Catchment phosphorus             |

: {.sm .hover}

In this analysis, I'm interested in finding out which areas of New Zealand have relatively high or low levels of contamination given relatively high or low livestock density. Which lends itself well to visualising using a bivariate colour scale.

Livestock density is, of course, only one of a number of interacting pressures determining the state of water quality. But this technique only stretches to exploring two variables at once... Trivariate maps would be almost impossible to interpret! I suggest exploring more than two variables at a time may require moving to a full modelling approach and looking at plots like [Partial Dependence Plots](https://scikit-learn.org/stable/modules/partial_dependence.html) (PDP) or [**SH**apley **A**dditive ex**P**lanations](https://towardsdatascience.com/using-shap-values-to-explain-how-your-machine-learning-model-works-732b3f40e137) (SHAP) values.

```{r}
#| code-summary: "Show libraries code"

library(tidyverse)
library(sf)
library(mapview)
library(koordinatr)
library(biscale)
library(cowplot)
library(patchwork)
library(ggtext)

custom_pal4 <- c(
  "1-1" = "#f2f2f2", # low x, low y #ffffff
  "2-1" = "#b9ddff",
  "3-1" = "#64bbff",
  "4-1" = "#0295ff", # high x, low y
  "1-2" = "#fedcb7",
  "2-2" = "#b8b8b8",
  "3-2" = "#6691bb",
  "4-2" = "#0069ba",
  "1-3" = "#ffb869",
  "2-3" = "#b89168",
  "3-3" = "#676767",
  "4-3" = "#003867",
  "1-4" = "#ff9400", # low x, high y
  "2-4" = "#bb6800",
  "3-4" = "#693800",
  "4-4" = "#000000" # high x, high y
)
```

## A primer on bivariate maps

```{r}
#| label: fig-pal
#| fig-cap: "Example of bivariate colour scale grid."
#| column: margin
#| echo: false

bi_pal(custom_pal4, dim = 4, flip_axes = TRUE) +
  theme_classic(base_size = 20) +
  theme(
    axis.text = element_blank(),  # Remove axis text
    axis.ticks = element_blank(),  # Remove axis ticks
    axis.line = element_blank()   # Remove axis lines
  )

```

Bivariate colour scale maps are thematic maps that use two different colour scales to simultaneously display two variables on the same map. This approach allows for the visualisation of the relationship between the two variables across a geographic space. Each colour on the map represents a unique combination of the two variables' values.

Typically, one colour gradient is represented on the x-axis and the other on the y-axis. The intersection of these gradients results in a grid of colours, each representing a different combination of the two variables' values (@fig-pal).

## Livestock vs water quality

### Data sources

The data for stocking density, *E. coli*, and nitrogen were obtained from the NIWA's latest report on spatial modelling of river water quality in New Zealand. This report utilised monitoring data from 2016 to 2020 to develop model-based predictions for approximately 590,000 unique river segments across the national river network.
[The study utilised random forest models to predict water quality states. This method involves using an ensemble of regression trees to improve prediction accuracy and handle complex, non-linear relationships between predictors and response variables.]{.aside}

The data used in this analysis includes:

-   **Stocking Density**: Represented as the number of livestock units per hectare. Derived from the Agricultural Production Survey hex grids.
-   **E. coli**: Modelled concentrations measured in colony-forming units per 100 millilitres (cfu/100ml). E. coli serves as an indicator of faecal contamination in water bodies.
-   **Nitrogen**: Modelled NNN concentrations measured in milligrams per litre (mg/L). Nitrogen levels are indicative of nutrient pollution, primarily from agricultural runoff.

```{r}
#| label: import
#| cache: true
#| output: false
#| code-summary: "Show data sources code"

# water quality
# ecoli_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109886")
# saveRDS(ecoli_raw, "data/ecoli_raw.rds")
ecoli_raw <- readRDS("data/ecoli_raw.rds")

ecoli <- ecoli_raw |>
  filter(mesrmnt == "Median") |>
  select(nzsgmnt,
    median_ecoli = value,
    strm_rd
  )

rm(ecoli_raw)

# nitrogen_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109888")
# saveRDS(nitrogen_raw, "data/nitrogen_raw.rds")
nitrogen_raw <- readRDS("data/nitrogen_raw.rds")

nitrogen <- nitrogen_raw |>
  filter(measr_b == "NNN") |>
  select(nzsgmnt,
    median_nitrogen = value
  )

rm(nitrogen_raw)

# load predictors
load("data/REC2.4_lcdb5_predictors.RData")
rm(ThesePredictors)

ecoli |>
  left_join(nitrogen |> st_drop_geometry(), by = join_by(nzsgmnt)) |>
  left_join(REC2.4_predictors, by = c("nzsgmnt" = "nzsegment")) |>
  select(
    nzsgmnt,
    median_ecoli,
    median_nitrogen,
    PropDairy_2017,
    PropSheep_2017,
    SUDensityTotal_2017,
    StreamOrder
  ) |>
  filter(StreamOrder > 1) |> 
  # filter extreme outliers 
  filter(SUDensityTotal_2017 < 40) -> dat
```

### Scatter plot

Let's start by looking at the relationship between total stock density and median *E. coli* levels. We can see that there is a positive relationship between the two variables - areas with stock density tend to have higher *E. coli* levels (@fig-scatter-plot).

However, the relationship is not perfect - there are areas with high stock density and low *E. coli* levels, and vice versa. This is where bivariate maps come in - they allow us to visualise the relationship between two variables across a geographic space.

```{r}
#| label: fig-scatter-plot
#| fig-cap: "Scatter plot of dairy density and median E. coli levels."

dat |> 
  ggplot(aes(SUDensityTotal_2017, median_ecoli)) +
  geom_point(alpha = 0.01, size = 1) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(x = "Dairy density", y = "Median E. coli")
```

### Univariate maps

Traditional maps use just a single colour scale to display a single variable (@fig-uni-maps). Displaying them side by side isn't a bad option - but it's pretty hard to make the comparison of locations with high dairy density and low *E. coli* (and vice versa) right? Well, that's where bivariate maps come in.

```{r}
#| label: fig-uni-maps
#| fig-cap: "Univariate maps of median E. coli levels and dairy density. The maps show the distribution of each variable across New Zealand, with high values represented in the traditional farming regions."
#| code-summary: "Show univariate code"

dat |>
  arrange(median_ecoli) |>
  ggplot() +
  geom_sf(aes(colour = median_ecoli)) +
  # scale_colour_viridis_c() +
  scale_colour_gradient(low = "#f2f2f2", high = "#0295ff") +
  theme_void(base_size = 8) +
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b"
  )) +
  labs(
    title = "E. coli concentrations",
    colour = "Modelled median \nconcentration of E. coli \n(cfu/100mL)"
  ) -> ecoli_map

dat |>
  ggplot() +
  geom_sf(aes(colour = SUDensityTotal_2017)) +
  scale_colour_gradient(low = "#f2f2f2", high = "#ff9400") +
  # scale_fill_viridis_c() +
  theme_void(base_size = 8) +
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b"
  )) +
  labs(
    title = "Total stock units",
    colour = "Catchment density \nof total stock units \n(SU/ha)"
  ) -> dairy_dens_map

ecoli_map + dairy_dens_map
```

### Bivariate colour scale maps

```{r}
#| label: spatial-join
#| code-summary: "Show prep code"

finalplot <- function(data, xlab = "Higher X", ylab = "Higher Y") {
  map <- ggplot() +
    geom_sf(
      data = data, mapping = aes(colour = bi_class), linewidth = 0.5, show.legend = FALSE,
      # color = "white"
    ) +
    bi_scale_color(pal = custom_pal4, dim = 4) +
    bi_theme()

  legend <- bi_legend(
    pal = custom_pal4,
    dim = 4,
    xlab = xlab,
    ylab = ylab,
    size = 7,
    pad_width = 0.5
  )

  finalPlot <- ggdraw() +
    draw_plot(map, 0, 0, 1, 1) +
    draw_plot(legend, 0.2, .65, 0.2, 0.2, 1.25)

  return(finalPlot)
}
  
```

#### Total livestock density

::: panel-tabset
## E. coli

```{r, dpi=300}
#| label: fig-bi-maps-total-ecoli

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = SUDensityTotal_2017,
  style = "fisher", dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher Stock"
  ) -> finalPlotTotalEcoli

finalPlotTotalEcoli
```

## Nitrogen

```{r, dpi=300}
#| label: fig-bi-maps-total-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = SUDensityTotal_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher stock"
  ) -> finalPlotTotalNitrogen

finalPlotTotalNitrogen
```
:::

#### Sheep density

::: panel-tabset
## E. coli

```{r}
#| label: fig-bi-maps-sheep-ecoli

bi_class(
  dat,
  x = median_ecoli,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher sheep"
  ) -> finalPlotSheepEcoli

finalPlotSheepEcoli
```

## Nitrogen

```{r}
#| label: fig-bi-maps-sheep-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher sheep"
  ) -> finalPlotSheepNitrogen

finalPlotSheepNitrogen
```
:::

#### Dairy density

::: panel-tabset
## E. coli

```{r}
#| label: fig-bi-maps-dairy-ecoli

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher dairy"
  ) -> finalPlotDairyEcoli

finalPlotDairyEcoli
```

## Nitrogen

```{r}
#| label: fig-bi-maps-dairy-nitrogen

bi_class(
  dat,
  x = median_nitrogen,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher dairy"
  ) -> finalPlotDairyNitrogen

finalPlotDairyNitrogen
```
:::

## Conclusion
