---
title: "Bivariate colour scales"
subtitle: "Mapping livestock density and water quality"
author: "Isaac Bain"
date: "2024-07-15"
categories: [code, water quality, maps]
execute:
  warning: false
  error: false
  messages: false
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: left
    number-sections: true
editor: visual
lightbox: auto
draft: true
title-block-banner: "#00000000"
title-block-banner-color: "rgba(255, 255, 255, 0.9)"
include-in-header:
  - text: |
      <style>
      #title-block-header.quarto-title-block.default .quarto-title-meta {
        color: rgba(255, 255, 255, 0.9);
      }
      .quarto-title-block .quarto-title-banner {
        height: 0; /* hide */
      }
      #title-block-header {
        background: 
          /* top, transparent black, faked with gradient */
          linear-gradient(
            rgba(0, 0, 0, 0.6),
            rgba(0, 0, 0, 0.2)
          ),
          /* bottom, image */ 
          url(./cows.jpg);
        background-size: cover;
        background-position-y: center;
        height: 300px;
        opacity: 0.7; /* image opacity, lower means lighter */
        z-index: -1;
      }
      </style>
---

## Overview

We all want rivers that are safe to swim in, clean water to drink, and healthy freshwater ecosystems. However, many areas in New Zealand are not meeting our communities' expectations and require improvement.[^1] Knowing where to target efforts to improve water quality is crucial, and a solid understanding of pressures is needed.

[^1]: <https://environment.govt.nz/publications/our-freshwater-2023/>

For management actions to be effective, they should be based on predictive relationships between land use, stressor levels in receiving waters, and the ecological and cultural responses to these stressors.[^2] Land use intensity is consistently identified as a key pressure on water quality in New Zealand. In both process-based and statistical models predicting water quality, land use is a significant predictor of *E. coli* and nitrogen concentrations.

[^2]: Scott T. Larned, Jonathan Moores, Jenni Gadd, Brenda Baillie & Marc Schallenberg (2020) Evidence for the effects of land use on freshwater ecosystems in New Zealand, New Zealand Journal of Marine and Freshwater Research, 54:3, 551-591, DOI: 10.1080/00288330.2019.1695634

For example, in NIWA's recent random forest modelling of river water quality[^3], the top ten most important predictors of nitrate + nitrite-nitrogen (NNN) and *E. coli* were:

[^3]: Snelder T. H., Fraser C., Larned S. T., Monaghan R., De Malmanche S., Whitehead A. L. (2022) Attribution of river water-quality trends to agricultural land use and climate variability in New Zealand. Marine and Freshwater Research 73, 1-19.

| Predictor rank | NNN predictors | *E. coli* predictors |
|-------------------|---------------------------|---------------------------|
| 1 | Stock density | Catchment elevation |
| 2 | Dairy density | Rainfall variability |
| 3 | Beef density | Stock density |
| 4 | Intensive agriculture land cover | Bare land cover |
| 5 | Urban land cover | Intensive agriculture land cover |
| 6 | Catchment slope | Slope |
| 7 | Temperature Minimum | Urban land cover |
| 8 | Rainfall variability | Exotic forest land cover |
| 9 | Number of warm days | Local elevation |
| 10 | Catchment elevation | Catchment phosphorus |

: {.sm .hover}

In this analysis, I aim to identify which areas of New Zealand have relatively high or low levels of contamination given relatively high or low livestock density. This is well-suited for visualisation using a bivariate colour scale.

Livestock density is one of many interacting pressures determining the state of water quality. This technique allows exploring two variables at once, but trivariate maps would be almost impossible to interpret! Exploring more than two variables at a time may require moving to a full modelling approach and looking at plots like [Partial Dependence Plots](https://scikit-learn.org/stable/modules/partial_dependence.html) (PDP) or [**SH**apley **A**dditive ex**P**lanations](https://towardsdatascience.com/using-shap-values-to-explain-how-your-machine-learning-model-works-732b3f40e137) (SHAP) values.

```{r}
#| code-summary: "Show libraries code"

library(tidyverse)
library(sf)
library(mapview)
library(koordinatr)
library(biscale)
library(cowplot)
library(patchwork)
library(ggtext)

custom_pal4 <- c(
  "1-1" = "#f2f2f2", # low x, low y #ffffff
  "2-1" = "#b9ddff",
  "3-1" = "#64bbff",
  "4-1" = "#0295ff", # high x, low y
  "1-2" = "#fedcb7",
  "2-2" = "#b8b8b8",
  "3-2" = "#6691bb",
  "4-2" = "#0069ba",
  "1-3" = "#ffb869",
  "2-3" = "#b89168",
  "3-3" = "#676767",
  "4-3" = "#003867",
  "1-4" = "#ff9400", # low x, high y
  "2-4" = "#bb6800",
  "3-4" = "#693800",
  "4-4" = "#000000" # high x, high y
)
```

## A Primer on Bivariate Maps

```{r}
#| label: fig-pal
#| fig-cap: "Example of bivariate colour scale grid."
#| column: margin
#| echo: false

bi_pal(custom_pal4, dim = 4, flip_axes = TRUE) +
  theme_classic(base_size = 20) +
  theme(
    axis.text = element_blank(),  # Remove axis text
    axis.ticks = element_blank(),  # Remove axis ticks
    axis.line = element_blank()   # Remove axis lines
  )

```

Bivariate colour scale maps are thematic maps that use two different colour scales to simultaneously display two variables on the same map. This approach allows for the visualisation of the relationship between the two variables across a geographic space. Each colour on the map represents a unique combination of the two variables' values.

Typically, one colour gradient is represented on the x-axis and the other on the y-axis. The intersection of these gradients results in a grid of colours, each representing a different combination of the two variables' values (@fig-pal).

## Livestock vs Water Quality

### Data sources

The data for stocking density, *E. coli*, and nitrogen (NNN) were obtained from the NIWA's latest report on spatial modelling of river water quality in New Zealand. This report utilised monitoring data from 2016 to 2020 to develop model-based predictions for approximately 590,000 unique river segments across the national river network. [The study utilised random forest models to predict water quality state. This method involves using an ensemble of regression trees to improve prediction accuracy and handle complex, non-linear relationships between predictors and response variables.]{.aside}

Selected data used in this analysis includes:

-   **Stocking Density**: Represented as the number of livestock units per hectare. Derived from the Agricultural Production Survey hex grids.
-   **E. coli**: Modelled concentrations measured in colony-forming units per 100 millilitres (cfu/100ml). *E. coli* serves as an indicator of faecal contamination in water bodies.
-   **Nitrogen**: Modelled NNN concentrations measured in milligrams per litre (mg/L). Nitrogen levels are indicative of nutrient pollution, primarily from agricultural runoff.

```{r}
#| label: import
#| cache: true
#| output: false
#| code-summary: "Show data code"

# water quality
# ecoli_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109886")
# saveRDS(ecoli_raw, "data/ecoli_raw.rds")
ecoli_raw <- readRDS("data/ecoli_raw.rds")

ecoli <- ecoli_raw |>
  filter(mesrmnt == "Median") |>
  select(nzsgmnt,
    median_ecoli = value,
    strm_rd
  )

rm(ecoli_raw)

# nitrogen_raw <- get_layer_as_sf(Sys.getenv("mfe_api_key"), "mfe", "109888")
# saveRDS(nitrogen_raw, "data/nitrogen_raw.rds")
nitrogen_raw <- readRDS("data/nitrogen_raw.rds")

nitrogen <- nitrogen_raw |>
  filter(measr_b == "NNN") |>
  select(nzsgmnt,
    median_nitrogen = value
  )

rm(nitrogen_raw)

# load predictors
load("data/REC2.4_lcdb5_predictors.RData")
rm(ThesePredictors)

ecoli |>
  left_join(nitrogen |> st_drop_geometry(), by = join_by(nzsgmnt)) |>
  left_join(REC2.4_predictors, by = c("nzsgmnt" = "nzsegment")) |>
  select(
    nzsgmnt,
    median_ecoli,
    median_nitrogen,
    PropDairy_2017,
    PropSheep_2017,
    SUDensityTotal_2017,
    StreamOrder
  ) |>
  filter(StreamOrder > 2) |> 
  # filter extreme outliers 
  filter(SUDensityTotal_2017 < 40) -> dat
```

### Scatter plot

Let's start by looking at the relationship between total stock density and median *E. coli* levels. There is a positive relationship between the two variables: areas with higher stock density tend to have higher *E. coli* levels (@fig-scatter-plot).

However, the relationship is not perfect. Some areas have high stock density and low *E. coli* levels, and vice versa. This is where bivariate maps come in, allowing us to visualise the relationship between two variables across a geographic space.

```{r}
#| label: fig-scatter-plot
#| fig-cap: "Scatter plot of total stock density and median E. coli levels."

dat |> 
  ggplot(aes(SUDensityTotal_2017, median_ecoli)) +
  geom_point(alpha = 0.01, size = 1) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(x = "Total stock density", y = "Median E. coli")
```

### Univariate maps

Traditional maps use just a single colour scale to display a single variable (@fig-uni-maps). Displaying them side by side isn't a bad option, but it's challenging but it’s challenging to compare locations with high stock density and low *E. coli* (and vice versa). Bivariate maps address this challenge.

```{r}
#| label: fig-uni-maps
#| fig-cap: "Univariate maps of median E. coli levels and dairy density."
#| code-summary: "Show univariate code"

dat |>
  arrange(median_ecoli) |>
  ggplot() +
  geom_sf(aes(colour = median_ecoli)) +
  # scale_colour_viridis_c() +
  scale_colour_gradient(low = "#f2f2f2", high = "#0295ff") +
  theme_void(base_size = 8) +
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b"
  )) +
  labs(
    title = "E. coli concentrations",
    colour = "Modelled median \nconcentration of E. coli \n(cfu/100mL)"
  ) -> ecoli_map

dat |>
  ggplot() +
  geom_sf(aes(colour = SUDensityTotal_2017)) +
  scale_colour_gradient(low = "#f2f2f2", high = "#ff9400") +
  # scale_fill_viridis_c() +
  theme_void(base_size = 8) +
  theme(plot.title = element_markdown(
    family = "sans", size = rel(1.5), face = "bold", color = "#4b015b"
  )) +
  labs(
    title = "Total stock units",
    colour = "Catchment density \nof total stock units \n(SU/ha)"
  ) -> dairy_dens_map

ecoli_map + dairy_dens_map
```

### Bivariate colour scale maps

In these bivariate maps, we use two intersecting colour scales to display two variables on a single map. The colour scale is divided into four quadrants, each representing a combination of high and low values of the two variables.

```{r}
#| label: spatial-join
#| code-summary: "Show prep code"

finalplot <- function(data, xlab = "Higher X", ylab = "Higher Y") {
  map <- ggplot() +
    geom_sf(
      data = data, mapping = aes(colour = bi_class), linewidth = 0.5, show.legend = FALSE,
      # color = "white"
    ) +
    bi_scale_color(pal = custom_pal4, dim = 4) +
    bi_theme()

  legend <- bi_legend(
    pal = custom_pal4,
    dim = 4,
    xlab = xlab,
    ylab = ylab,
    size = 7,
    pad_width = 0.5
  )

  finalPlot <- ggdraw() +
    draw_plot(map, 0, 0, 1, 1) +
    draw_plot(legend, 0.2, .65, 0.2, 0.2, 1.25)

  return(finalPlot)
}
  
```

#### Total livestock density

Remembering from above that total livestock density is "the catchment density of total stock units (SU) on pastoral land", i.e. the density of all the livestock types on all the land upstream of the relevant river segment, where the water quality is also modelled.

::: panel-tabset
## E. coli

Key findings:

-   Large areas of the Waikato have both high stock density and high *E. coli* levels.
-   Cities like Auckland, Wellington, and Christchurch have low stock density but high *E. coli* levels.
-   Hawke's Bay has relatively high stock density but low *E. coli* levels.

```{r}
#| label: fig-bi-maps-total-ecoli
#| fig-cap: "Bivariate maps of median E. coli levels and total stock density."

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = SUDensityTotal_2017,
  style = "fisher", dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher Stock"
  )

```

## Nitrogen

Key findings:

-   Canterbury and Southland have high nitrogen levels and high stock density, although some areas have high nitrogen levels and low stock density.
-   Northland, Hawke's Bay, and eastern Wairarapa have low nitrogen levels and high stock density.
-   Waikato and Taranaki have high stock densities but relatively lower nitrogen levels compared to Canterbury and Southland.

```{r}
#| label: fig-bi-maps-total-nitrogen
#| fig-cap: "Bivariate maps of median nitrogen levels and total stock density."

bi_class(
  dat,
  x = median_nitrogen,
  y = SUDensityTotal_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher stock"
  ) -> finalPlotTotalNitrogen

finalPlotTotalNitrogen
```
:::

#### Sheep density

We can also separate the total stock density predictor into specific livestock types to see if the patterns differ for different types of livestock. Here, we show the results for sheep density.

::: panel-tabset
## E. coli

Key findings:

-   Waikato and Northland have low sheep density for relatively high *E. coli* levels.
-   The Canterbury high country, and Otago have high sheep density for relatively low *E. coli* levels.
-   The Canterbury plains have low sheep density for a relatively high *E. coli* levels.

```{r}
#| label: fig-bi-maps-sheep-ecoli
#| fig-cap: "Bivariate maps of median E. coli levels and sheep density."

bi_class(
  dat,
  x = median_ecoli,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher sheep"
  ) -> finalPlotSheepEcoli

finalPlotSheepEcoli
```

## Nitrogen

Key findings:

-   Manawatū-Whanganui, Hawke's Bay, Wairarapa, Marlborough, Canterbury high country, and Otago have high sheep density but relatively low levels of nitrogen.
-   The Canterbury plains, and to a lesser extent Waikato, have low sheep density but relatively high levels of nitrogen.

```{r}
#| label: fig-bi-maps-sheep-nitrogen
#| fig-cap: "Bivariate maps of median nitrogen levels and sheep density."

bi_class(
  dat,
  x = median_nitrogen,
  y = PropSheep_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher sheep"
  ) -> finalPlotSheepNitrogen

finalPlotSheepNitrogen
```
:::

#### Dairy density

Here's a similar analysis for dairy density.

::: panel-tabset
## E. coli

Key findings:

-   Some areas of Northland, Waikato, Taranaki, Manawatū-Whanganui, and Southland have both high levels of dairy density and *E. coli*.
-   Canterbury has high levels of dairy density throughout, but lower levels of *E. coli* in some rivers, presumably due to the large alpine-fed systems.
-   Bay of Plenty and the West Coast have relatively lower *E. coli* levels for a given dairy density.

```{r}
#| label: fig-bi-maps-dairy-ecoli
#| fig-cap: "Bivariate maps of median E. coli levels and dairy density."

# create classes
bi_class(
  dat,
  x = median_ecoli,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher E.coli",
    ylab = "Higher dairy"
  ) -> finalPlotDairyEcoli

finalPlotDairyEcoli
```

## Nitrogen

Key findings:

-   Canterbury, Southland, and the lower Waikato have both high dairy density and nitrogen levels.
-   Hawke's Bay and eastern Wairarapa have medium level nitrogen levels despite having low dairy density.

```{r}
#| label: fig-bi-maps-dairy-nitrogen
#| fig-cap: "Bivariate maps of median nitrogen levels and dairy density."

bi_class(
  dat,
  x = median_nitrogen,
  y = PropDairy_2017,
  style = "fisher",
  dim = 4
) |>
  finalplot(
    xlab = "Higher nitrogen",
    ylab = "Higher dairy"
  ) -> finalPlotDairyNitrogen

finalPlotDairyNitrogen
```
:::

## Conclusions

This analysis highlights the complex relationship between livestock density and water quality in New Zealand. Using bivariate colour scale maps, we can visually explore these interactions and identify regions where specific patterns emerge.